Robert C. Dunnell’s _Systematics in Prehistory_ at 50

### 2022-04-20 analysis/paper/paper.Rmd
update text around acceptance

---
title: "Robert C. Dunnell’s _Systematics in Prehistory_ at 50"
author:
  - Felix Riede:
      email: f.riede@cas.au.dk
      institute: [UofO]
      correspondence: true
  - Astolfo Araujo:
      institute: [UofSP]
      correspondence: false
  - Ben Marwick:
      email: bmarwick@uw.edu
      institute: [UofWA]
      correspondence: false
institute:
  - UofO: Department of Archaeology and Heritage Studies, Aarhus University, Moesgård Allé 20, 8270 Højbjerg, Denmark
  - UofSP: Museum of Archaeology and Ethnology, University of São Paulo, Brazil
  - UofWA: Department of Anthropology, University of Washington, Seattle, WA, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/evolutionary-anthropology.csl" # Insert path for the bib-sty
abstract: |
  2021 marks the fiftieth anniversary of the publication of Robert C. Dunnell’s (1971) diminutive yet dense _Systematics in Prehistory_. At the height of the debate between Culture History and New Archaeology, Dunnell’s work sought to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Paleolithic: systematics. Dunnell himself was notorious and controversial, however, but the importance of his work remains underappreciated. Like other precocious works of that tumultuous time Systematics in Prehistory today remains absent from most course reading lists and gathers dust on library shelves. In this contribution we argue for a greater appreciation of its as yet unfulfilled conceptual and analytical promise. In particular, we briefly chart its somewhat delayed impact via evolutionary archaeology, including how it has also influenced non-Anglophone traditions, especially in South America. The obstinate persistence of classification issues in paleoanthropology and Paleolithic archaeology, we argue, warrants a second look at Dunnell's _Systematics_.
---

@@

We found `r how_many_citing_works` citations of *Systematics*; for reference the most highly cited archaeology publication from the same year is Binford's [@binford1971mortuary] 'Mortuary Practices: Their Study and Their Potential' with 1621 citations currently on Google Scholar at the time of our data collection. Citations to *Systematics* have steadily accumulated over time, with a distinct increase in the annual rate of citation in the early 2000s as evolutionary concepts become increasingly integrated into archaeological science (Figure \@ref(fig:sip-citations-of-citing-work)). Several works that cite *Systematics* have themselves been very highly cited, with ten publications receiving over 500 citations, demonstrating its influence over a range of topics including social theory [@Shanks_Tilley_1987], behavioral archaeology [@schiffer2016behavioral], and lithic analysis [@Andrefsky_1998]. Looking at the top fifty words in the titles of the works citing *Systematics*, we can also see that 'lithics' and 'stone' are prominent, indicating that its contribution was especially noted by archaeologists working on stone artefacts (Figure \@ref(fig:sip-keywords-in-titles)). Pottery is the only other artefact type where analysts substantially engaged with *Systematics*. Notable in the title keywords are concepts relating to evolution, for example, 'transmission', 'variation', and 'evolutionary'. Evolution is a very minor theme in *Systematics* itself, but these word frequency data show it has proven to be a foundational text in applications of evolutionary theory to explaining variability in the archaeological record.

@@

Beyond the Anglosphere, we see citations from a small number works in Russian and European languages (Figure \@ref(fig:sip-keywords-in-titles)), but the impact of *Systematics* is most evident in Spanish-language publications by scholars working on South American stone artefact assemblages. These South American studies have been especially innovative, taking to heart the critique in *Systematics* that archaeological typologies are often intuitive, arbitrary, and difficult to replicate by other researchers. Motivated by this critique, scholars such as Marcelo Cardillo, Judith Charlin [@Cardillo_Charlin_2018] and others have conducted pioneering work in the application of geometric morphometrics to stone artefact assemblages in an effort to provide a materialist view of technological variation where the focus is on continuous quantitative phenomena. While geometric morphometrics has been applied by archaeologists to a range of regions and artefact types (including ceramic and metal), what makes this South American work remarkable as part of the legacy of *Systematics* is their exploration of modern phylogenetic comparative methods to model and quantify technological variation and change over space and time [e.g. @cardillo2015evolution]. These works show that one of the most important contributions of *Systematics* has been as a bridge between archaeology and cognate fields - first and foremost palaeontology - dealing with challenges of classification and concerned with modelling macroevolutionary processes. The bridge-building that has followed from this has been very fruitful, reflected, for example, in the increasing number of archaeological presentations at meetings of the Cultural Evolution Society since the first one in 2017.

@@

A second limitation of *Systematics* is that it failed to motivate the formation of a distinct research area and community of archaeological systematics. While biology has a Society of Systematic Biologists, and attendant journals such as *Systematic Biology*, *Systematics and Evolution*, and *Systematics and Biodiversity*, archaeology currently has no equivalent focal points for discussions of systematics that transcends specific geographical and chronological concerns. *Systematics* was not a work that served as an engine for moving archaeologists together at scale to tackle questions about the formation of units of measurement and classification. Instead, these discussions typically happen deep within discrete and disconnected archaeological research communities, constrained by the culture and norms of those groups. The result of this fragmentation is a proliferation of bespoke classifications that lack applicability across cases. It also inhibits reproducibility, synthesis, and the transfer of ideas and innovations across research communities, which may limit the sustainability of archaeology as a discipline.

>>>>

---
title: "Robert C. Dunnell’s _Systematics in Prehistory_ at 50"
author:
  - Felix Riede:
      email: f.riede@cas.au.dk
      institute: [UofO]
      correspondence: true
  - Astolfo Araujo:
      institute: [UofSP]
      correspondence: false
  - Ben Marwick:
      email: bmarwick@uw.edu
      institute: [UofWA]
      correspondence: false
institute:
  - UofO: Department of Archaeology and Heritage Studies, Aarhus University, Moesgård Allé 20, 8270 Højbjerg, Denmark
  - UofSP: Museum of Archaeology and Ethnology, University of São Paulo, Brazil
  - UofWA: Department of Anthropology, University of Washington, Seattle, WA, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
       - --lua-filter=../templates/scholarly-metadata.lua
       - --lua-filter=../templates/author-info-blocks.lua
       - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/apa.csl" # Insert path for the bib-sty
abstract: |
  2021 marks the fiftieth anniversary of the publication of Robert C. Dunnell’s (1971) diminutive yet dense _Systematics in Prehistory_. At the height of the debate between Culture History and New Archaeology, Dunnell’s work sought to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Paleolithic: systematics. Dunnell himself was notorious and controversial, however, but the importance of his work remains underappreciated. Like other precocious works of that tumultuous time Systematics in Prehistory today remains absent from most course reading lists and gathers dust on library shelves. In this contribution we argue for a greater appreciation of its as yet unfulfilled conceptual and analytical promise. In particular, we briefly chart its somewhat delayed impact via evolutionary archaeology, including how it has also influenced non-Anglophone traditions, especially in South America. The obstinate persistence of classification issues in paleoanthropology and Paleolithic archaeology, we argue, warrants a second look at Dunnell's _Systematics_.
---

@@

We found `r how_many_citing_works` citations of *Systematics*; for reference the most highly cited archaeology publication from the same year is Binford's [-@binford1971mortuary] 'Mortuary Practices: Their Study and Their Potential' with 1621 citations currently on Google Scholar at the time of our data collection. Citations to *Systematics* have steadily accumulated over time, with a distinct increase in the annual rate of citation in the early 2000s as evolutionary concepts become increasingly integrated into archaeological science (Figure \@ref(fig:sip-citations-of-citing-work)). Several works that cite *Systematics* have themselves been very highly cited, with ten publications receiving over 500 citations, demonstrating its influence over a range of topics including social theory [@Shanks_Tilley_1987], behavioral archaeology [@schiffer2016behavioral], and lithic analysis [@Andrefsky_1998]. Looking at the top fifty words in the titles of the works citing *Systematics*, we can also see that 'lithics' and 'stone' are prominent, indicating that its contribution was especially noted by archaeologists working on stone artefacts (Figure \@ref(fig:sip-keywords-in-titles)). Pottery is the only other artefact type where analysts substantially engaged with *Systematics*. Notable in the title keywords are concepts relating to evolution, for example, 'transmission', 'variation', and 'evolutionary'. Evolution is a very minor theme in *Systematics* itself, but these word frequency data show it has proven to be a foundational text in applications of evolutionary theory to explaining variability in the archaeological record.

@@

Beyond the Anglosphere, we see citations from a small number works in Russian and European languages (Figure \@ref(fig:sip-keywords-in-titles)), but the impact of *Systematics* is most evident in Spanish-language publications by scholars working on South American stone artefact assemblages. These South American studies have been especially innovative, taking to heart the critique in *Systematics* that archaeological typologies are often intuitive, arbitrary, and difficult to replicate by other researchers. Motivated by this critique, scholars such as Marcelo Cardillo, Judith Charlin [@Cardillo_Charlin_2018], Mercedes Okumura [@Okumura_Araujo_2014] and others have conducted pioneering work in the application of geometric morphometrics to stone artefact assemblages in an effort to provide a materialist view of technological variation where the focus is on continuous quantitative phenomena. While geometric morphometrics has been applied by archaeologists to a range of regions and artefact types (including ceramic and metal), what makes this South American work remarkable as part of the legacy of *Systematics* is their exploration of modern phylogenetic comparative methods to model and quantify technological variation and change over space and time [e.g. @cardillo2015evolution]. While the work by Cardillo and colleagues [-@Cardillo_Charlin_2018; -@cardillo2015evolution] focuses on phylogenetic signals and material culture diversity across time and space, Okumura and colleagues [-@Okumura_Araujo_2014] are more concerned with the persistence of attributes over long stretches of time. What these works have in common is the use of an explicitly cultural evolutionary rationale. In these studies, the use of paradigmatic classifications is widespread and an overt understanding of the difference between classes (ideational/theoretical) and groups (phenomenological/empirical) evident, underlining the influence of Dunnell’s thinking. Worth noting is that these approaches were embraced chiefly in only two South American countries – Argentina and Brazil – and largely independently of each other. Collectively, these works show that one of the most important contributions of *Systematics* has been as a bridge between archaeology and cognate fields - first and foremost palaeontology - dealing with challenges of classification and concerned with modelling macroevolutionary processes. The bridge-building that has followed from this has been very fruitful, reflected, for example, in the increasing number of archaeological presentations at meetings of the Cultural Evolution Society since the first one in 2017.

@@

A second limitation of *Systematics* is that it failed to motivate the formation of a distinct research area and community of archaeological systematics. While biology has a Society of Systematic Biologists, and attendant journals such as *Systematic Biology*, *Systematics and Evolution*, and *Systematics and Biodiversity*, archaeology currently has no equivalent focal points for discussions of specifically culture-historical systematics that transcend specific geographical and chronological concerns. *Systematics* was not a work that served as an engine for moving archaeologists together at scale to tackle questions about the formation of units of measurement and classification. Instead, these discussions typically happen deep within discrete and disconnected archaeological research communities, constrained by the culture and norms of those groups. The result of this fragmentation is a proliferation of bespoke classifications that lack applicability across cases. It also inhibits reproducibility, synthesis, and the transfer of ideas and innovations across research communities, which may limit the sustainability of archaeology as a discipline.
